Automatic Parallelization and Transparent Fault Tolerance\\ (\emph{New project} paper)

### 2016-04-19 pubs/tfp16/paper.tex
random edits

\subsection{STG to C}
The back-end is written in Haskell and generates C code; the runtime system is
written in C/C++. This gives us the extensive optimization performed by modern
C compilers such as gcc and Clang/LLVM. The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which may be different than those for non-strict semantics.
Sharing is preserved.

\subsubsection{Proper tail calls in C.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel
execution they identified the need for the ability to examine and unwind the
stack in quashing concurrent evaluation of the same \emph{THUNK}\@.  This
requires that use of the C stack be highly constrained if we are to avoid
complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
code to directly manipulate the C stack, complicating an already non-trivial
implementation effort.  We therefore must limit the use of the C stack to
calls to the runtime, in other words, when executing code generated from the
user program (and not in a runtime call), there must be no growing of (pushing
return addresses or data onto) the C stack.
%
To clearly distinguish between a notional C call (a syntactic concept) and a
machine-level call (pushing a return address onto the native C stack) we will
refer to the latter as \emph{call-with-return}, and to a machine-level tail
call as \emph{jump}.

First we establish that this is possible in principle.  By maintaining our own
data and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls-with-return for C
calls.  When using gcc we can also use a fixed set of machine registers for
passing/returning values, much like GHC using the LLVM back-end with its
custom GHC calling convention.

The STG abstract machine is implemented using a continuation stack.
Continuations contain a code pointer, optionally some data, and layout
information for the benefit of the garbage collector.  For example, the
argument frame for a function call is one type of continuation.  When the
evaluation of a C function (corresponding to some STG code) is completed its
return is by an indirect jump using the code pointer in the continuation on
the top of the stack (``returning through the continuation stack'').  In the
case of an argument frame for a function, the associated code simply pops the
continuation off the stack and returns through the next continuation on the
stack.

@@

In addition to being in a tail call position, the gcc constraints for
generating a tail call (jump) are (approximately) that the total size of the
caller and callee arguments be equal, and the return type sizes be equal.  For
Clang the requirement is slightly more strict: caller and callee type
signatures must be the same.  These are a consequence of the C calling
convention that the caller, not callee, cleans up the stack (removes the
callee's stack frame).  However, because we maintain our own stack we can
define our own calling convention(s), so tail calls can be made to functions
of notionally differing type.  In particular, top-of-stack frames can be
adjusted in size and content without undue copying.
\end{comment}

@@

We are currently developing the parallel runtime.  Parallel garbage collection
is likely out of scope.
%
The fault-tolerance mechanisms remain future work.

@@

Funding for this project has been provided by the DOE NNSA LANL Laboratory
Directed Research and Development program award 20150845ER; the National
Science Foundation Science, Technology, Engineering, and Mathematics Talent
Expansion Program (STEP) program for undergraduate students; and the DOE
Science Undergraduate Laboratory Internship (SULI) program.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy's National Nuclear Security Administration (NNSA).

@@

\subsubsection{Avoiding building thunks.}
As \emph{EA} points out, there is no need to build a thunk for the scrutinee
of a \texttt{case} expression because if the \texttt{case} is evaluated the
scrutinee is certain to be evaluated.  Similarly, if a function is strict
there is no need to build a thunk for each of its arguments (assuming boxed
argument type).  Because STG allows only \emph{atoms} (variables or literal
values) in argument positions, naive translation of Haskell \texttt{f e} to
\texttt{let \{x = e\} in f x}, followed by naive code generation, will result in
an unnecessary heap allocation.  An alternative is to translate to
\texttt{case e of x \{\_ -> f x\}}.

@@

\begin{verbatim}
append = FUN(l1 l2 ->
            case l1 of
              { Nil -> l2;
                Cons hd tl -> 
                  let { rec = THUNK(append tl l2);
                        result = CON(Cons hd rec) }
                  in result });
\end{verbatim}
In mini-Haskell, expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
\begin{verbatim}
append l1 l2 = case l1 of
                 Nil -> l2
                 Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}
This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

Even relatively simple functions can be painful to write, such as \texttt{unzip}.
\begin{verbatim}
unzip = FUN(list ->
  case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
                 Cons x xs ->
                   case x of {
                     TP2 a b ->     --TP2 is the Tuple constructor
                       case unzip xs of {
                         TP2 as bs ->
                           let {aas = CON(Cons a as);
                                bbs = CON(Cons b bs);
                                res = CON(TP2 aas bbs)}
                           in res; }}});
\end{verbatim}
In mini-Haskell, this becomes
\begin{verbatim}
unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
unzip Nil = TP2 Nil Nil
unzip (Cons (TP2 a b) xs) =
  case unzip xs of
    TP2 as bs -> TP2 (Cons a as) (Cons b bs)
\end{verbatim}

>>>>

\subsection{STG to C}
The back-end is written in Haskell and generates C code; the runtime system is
written in C/C++. This gives us the extensive optimization performed by modern
C compilers such as GCC and Clang/LLVM. The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which may be different than those for non-strict semantics.
Sharing is preserved.

\subsubsection{Proper tail calls in C.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel
execution they identified the need for the ability to examine and unwind the
stack in quashing concurrent evaluation of the same \emph{THUNK}\@.  This
requires that use of the C stack be highly constrained if we are to avoid
complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
code to directly manipulate the C stack, complicating an already non-trivial
implementation effort.  We therefore must limit the use of the C stack to
calls to the runtime, in other words, when executing code generated from the
user program (and not in a runtime call), there must be no growing of (pushing
return addresses or data onto) the C stack.
%
To clearly distinguish what a C call (a syntactic concept) can compile to, we
will refer to a machine-level call (pushing a return address onto the native C stack)
as \emph{call-with-return}, and to a machine-level jump as \emph{jump}.

First we establish that this is possible in principle on x86-64/Linux.  By
maintaining our own data and control (continuation) stack all generated C
functions can take a uniform (including empty) set of explicit arguments and
have uniform return type). If we arrange that generated C code has only calls
(other than runtime or other external calls, which do not call back to
user-program code) in tail call positions, that is, immediately before a
return in the control flow graph, then at non-zero optimization levels both
GCC and Clang/LLVM will generate jumps instead of calls-with-return for C
calls.  When using GCC we can also use an explicit set of machine registers
for passing/returning values, much like GHC using the LLVM back-end with its
custom GHC calling convention.

The STG abstract machine is implemented using a combined data and continuation stack.
Continuations contain a code pointer, optionally some data, and layout
information for the benefit of the garbage collector.  
% For example, the argument frame for a function call is one type of
% continuation.
When the evaluation of a C function (corresponding to some STG code) is
completed its return is by an indirect jump using the code pointer in the
continuation on the top of the stack (``returning through the continuation
stack'').
%In the case of an argument frame for a function, the associated code simply
%pops the continuation off the stack and returns through the next continuation
%on the stack.

@@

In addition to being in a tail call position, the GCC constraints for
generating a tail call (jump) are (approximately) that the total size of the
caller and callee arguments be equal, and the return type sizes be equal.  For
Clang the requirement is slightly more strict: caller and callee type
signatures must be the same.  These are a consequence of the C calling
convention that the caller, not callee, cleans up the stack (removes the
callee's stack frame).  However, because we maintain our own stack we can
define our own calling convention(s), so tail calls can be made to functions
of notionally differing type.  In particular, top-of-stack frames can be
adjusted in size and content without undue copying.
\end{comment}

@@

For simplicity the implementation is 64-bit only.
%
The current prototype parallel runtime is based on Pthreads and a non-blocking
work queue~\cite{Michael:1996}.  We are currently developing a more realistic
parallel runtime, experimenting with using the Argobots user-level thread library~\cite{Seo:2015}.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.

@@

Funding for this project has been provided by the DOE NNSA LANL Laboratory
Directed Research and Development program award 20150845ER; the National
Science Foundation Science, Technology, Engineering, and Mathematics Talent
Expansion Program (NSF STEP) program for undergraduate students; and the Department of Energy
Science Undergraduate Laboratory Internship (DOE SULI) program.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy's National Nuclear Security Administration (NNSA).

@@

\subsubsection{Avoiding building thunks.}
As \emph{EA} points out, there is no need to build a thunk for the scrutinee
of a \texttt{case} expression because if the \texttt{case} is evaluated the
scrutinee is certain to be evaluated.  Similarly, if a function is strict
there is no need to build a thunk for each of its arguments (assuming boxed
argument type).  Because STG allows only \emph{atoms} (variables or literal
values) in argument positions, naive translation of Haskell \texttt{f e} to
\texttt{let \{x = e\} in f x}, followed by naive code generation, will result in
an unnecessary heap allocation.  An alternative is to translate to
\texttt{case e of x \{\_ -> f x\}}, which we regard as a role of a front-end
such as mini-Haskell.

@@

\begin{verbatim}
append = FUN(l1 l2 ->
            case l1 of
              { Nil -> l2;
                Cons hd tl -> 
                  let { rec = THUNK(append tl l2);
                        result = CON(Cons hd rec) }
                  in result })
\end{verbatim}
In mini-Haskell, expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
\begin{verbatim}
append l1 l2 = case l1 of
                 Nil -> l2
                 Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}
This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

Even relatively simple functions can be painful to write, such as \texttt{unzip}.
\begin{verbatim}
unzip = FUN(list ->
  case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
                 Cons x xs ->
                   case x of {
                     TP2 a b ->     --TP2 is the Tuple constructor
                       case unzip xs of {
                         TP2 as bs ->
                           let {aas = CON(Cons a as);
                                bbs = CON(Cons b bs);
                                res = CON(TP2 aas bbs)}
                           in res }}})
\end{verbatim}
In mini-Haskell, this becomes
\begin{verbatim}
unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
unzip Nil = TP2 Nil Nil
unzip (Cons (TP2 a b) xs) =
  case unzip xs of
    TP2 as bs -> TP2 (Cons a as) (Cons b bs)
\end{verbatim}
